### 2024-02-13 paper/paper.qmd
introduction added

---
title: "Racial Force Probe"
subtitle: "The study looks on a single scientific argument about whether or not certain police officers discriminate against people based on their race. "
author:
  - Monica Sainani
  - Kaavya Kalani
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction
Health and Safety of Prisoners Today**

Individuals' health and safety in jail are critical aspects of the criminal justice system
that must be thoroughly examined. As we negotiate the complicated world of prisons in contemporary America,
it becomes clear that the physical and emotional health of jailed people is a major problem. This study aims to give a
complete analysis of the health difficulties encountered by inmates today, diving into the areas of physical health,
emotional well-being, death rates, and the subtle distinctions between male and female prisoners.

Physical Health of Prisoners: Unveiling a Disturbing Reality

The first aspect of our investigation is a thorough examination of the physical health of convicts.
Recent discoveries in America portray a troubling picture, demonstrating that a sizable proportion of offenders enter
correctional facilities with a variety of pre-existing physical health issues.
The jail environment, which is marked by overcrowding, restricted access to healthcare, and a lack of preventive
measures, frequently exacerbates these problems. According to studies, the prison population has a high prevalence of
chronic diseases such as diabetes, hypertension, and infectious diseases, emphasizing the critical need to focus on
the physical well-being of jailed people.

The sad fact is that convicts have far lower health outcomes than the general population.
Access to medical care in prisons remains a recurrent issue, resulting in delayed diagnoses and insufficient treatment.
The cumulative effect of these conditions adds to a cycle of declining physical health among inmates, raising not just
ethical problems but also public health dangers when they are eventually released into society.

Mental Health: The Silent Crisis Behind Bars

Moving beyond physical health, the second half of our investigation focuses on the sometimes-ignored topic of mental
well-being in the prison system. Mental health difficulties among inmates have reached crisis proportions, with a slew
of obstacles compounding the psychological toll of incarceration. Suicide rates among convicts have become a major
problem, underlining the critical need for mental health services in correctional facilities.

Substance addiction exacerbates the mental health problem since many inmates take drugs to cope with the stresses of
confinement. The absence of comprehensive addiction treatment services in jails reinforces the cycle of substance misuse,
exacerbating mental health issues and impeding successful recovery attempts.

The mental health crisis among inmates emphasizes the significance of treating the underlying causes of psychological
anguish in correctional facilities. Initiatives to provide mental health support, provide counselling services, and
establish drug addiction treatment programs are critical to interrupting the cycle of mental health worsening among inmates.

From Suicides to Mortality Rates: Connecting the Dots

The third feature of our investigation focuses on the relationship between suicides and prison death rates.
Suicides in correctional facilities are not isolated instances; they are inextricably linked to larger concerns such as
mental health, poor healthcare, and the harsh reality of incarceration. Beyond suicides, inmates are always at risk of
death from a variety of causes, including murders, infectious infections, and other health difficulties.

Understanding the interconnectedness of these mortality variables is critical for designing holistic measures to improve
prisoner safety and well-being. It demands a comprehensive strategy that tackles both immediate concerns, such as
suicide prevention, as well as the larger structural factors that contribute to higher prison death rates.

**Gender Disparities in Prison Health: Unveiling Unique Challenges**

In the third portion of our investigation, we examine the subtle disparities in the health experiences of male and female
convicts. While many health issues are common, gender-specific characteristics play an important role in defining the health
outcomes of jailed people. Women inmates, for example, frequently confront particular reproductive health challenges,
as well as an increased risk of sexual assault while in correctional facilities.

Understanding and treating gender differences in prison health is critical for establishing focused therapies that
meet the unique requirements of both male and female inmates. Recognizing the unique issues experienced by each gender
group is an important step toward establishing a more equal and compassionate approach to healthcare in the prison system.

**Conclusion: A Call for Holistic Reforms**

To summarize, the health and safety of inmates today require immediate attention
and thorough improvements. The complex network of physical health issues, mental health crises,
mortality risks, and gender imbalances inside the jail system needs a comprehensive strategy.
Addressing these concerns necessitates not just urgent actions, such as increased healthcare access and mental health
assistance, but also a larger commitment to rethinking the purpose and consequences of incarceration on individual
well-being. This article intends to add to the continuing discussion about prison health by arguing for policies and
practices that prioritize the dignity, safety, and general well-being of people housed in correctional institutions.

# Data {#sec-data}

# Model

# Results

# Discussion

## Weaknesses and next steps

# Appendix {-}

# Additional data details
# References

### 2024-02-13 paper/paper.qmd
introduction added with citations

@@
 # Introduction
-Health and Safety of Prisoners Today**
+Health and Safety of Prisoners Today
 
@@
 The jail environment, which is marked by overcrowding, restricted access to healthcare, and a lack of preventive
-measures, frequently exacerbates these problems. According to studies, the prison population has a high prevalence of
+measures, frequently exacerbates these problems(Stevenson, B. (2021)). According to studies, the prison population has a high prevalence of
 chronic diseases such as diabetes, hypertension, and infectious diseases, emphasizing the critical need to focus on
@@
 The cumulative effect of these conditions adds to a cycle of declining physical health among inmates, raising not just
-ethical problems but also public health dangers when they are eventually released into society.
+ethical problems but also public health dangers when they are eventually released into society(M. Maruschak, L., & Berzofsky, M. (2015)).
 
@@
 well-being in the prison system. Mental health difficulties among inmates have reached crisis proportions, with a slew
-of obstacles compounding the psychological toll of incarceration. Suicide rates among convicts have become a major
+of obstacles compounding the psychological toll of incarceration.(Edgemon, T. G., & Clay-Warner, J. (2019)). Suicide rates among convicts have become a major
 problem, underlining the critical need for mental health services in correctional facilities.
@@
 confinement. The absence of comprehensive addiction treatment services in jails reinforces the cycle of substance misuse,
-exacerbating mental health issues and impeding successful recovery attempts.
+exacerbating mental health issues and impeding successful recovery attempts.(Edgemon, T. G., & Clay-Warner, J. (2019))
 
@@
 mental health, poor healthcare, and the harsh reality of incarceration. Beyond suicides, inmates are always at risk of
-death from a variety of causes, including murders, infectious infections, and other health difficulties.
+death from a variety of causes, including murders, infectious infections, and other health difficulties.(Emily Widra, (2017))
 
 Understanding the interconnectedness of these mortality variables is critical for designing holistic measures to improve
-prisoner safety and well-being. It demands a comprehensive strategy that tackles both immediate concerns, such as
+prisoner safety and well-being((Emily Widra, (2017)). It demands a comprehensive strategy that tackles both immediate concerns, such as
 suicide prevention, as well as the larger structural factors that contribute to higher prison death rates.
 
-**Gender Disparities in Prison Health: Unveiling Unique Challenges**
+Gender Disparities in Prison Health: Unveiling Unique Challenges
 
@@
 
-**Conclusion: A Call for Holistic Reforms**
+Conclusion: A Call for Holistic Reforms
 
@@
 assistance, but also a larger commitment to rethinking the purpose and consequences of incarceration on individual
-well-being. This article intends to add to the continuing discussion about prison health by arguing for policies and
+well-being(Ghazanfari, H., & Farokhzadian, J. (2022)).This article intends to add to the continuing discussion about prison health by arguing for policies and
 practices that prioritize the dignity, safety, and general well-being of people housed in correctional institutions.

### 2024-02-13 paper/paper.qmd
abstract added

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations.
+The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being.
+Utilizing data collected over the past [insert number of years], this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners.
+By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being.
+This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities.
+Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
 format: pdf
+toc: true
 number-sections: true
@@
 # Additional data details
+
 # References

### 2024-02-14 scripts/99-replication.R
Replication completed

@@ -1,0 +1,162 @@
+#### Preamble ####
+# Purpose: Replicated graphs from https://pubs.aeaweb.org/doi/pdfplus/10.1257/jep.35.4.97
+# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
+# Date: 12 February 2024
+# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+# Note: I will be replicating Figure 2, Figure 3 and Figure 4 from the paper
+
+
+#### Workspace setup ####
+# install.packages("tidyverse")
+# install.packages("gdata")
+# install.packages("dplyr")
+# install.packages("haven")
+# install.packages("ggplot2")
+# install.packages("zoo")
+library(tidyverse)
+library(gdata)
+library(dplyr)
+library(haven)
+library(ggplot2)
+library(zoo)
+
+# Figure 2 replication
+
+x <- read_dta("../data/spi86to16.dta")
+x <- x[x$admit == 1, ]
+
+# Create data frames for each program
+drugev <- x %>% group_by(region, year) %>% summarize(xbar = weighted.mean(drugev, wt, na.rm = TRUE))
+vocev <- x %>% group_by(region, year) %>% summarize(xbar = weighted.mean(vocev, wt, na.rm = TRUE))
+work <- x %>% group_by(region, year) %>% summarize(xbar = weighted.mean(work, wt, na.rm = TRUE))
+edev <- x %>% group_by(region, year) %>% summarize(xbar = weighted.mean(edev, wt, na.rm = TRUE))
+
+# Combine all data frames into a single data frame
+programs_data <- bind_rows(
+  mutate(drugev, program = "Drug Program"),
+  mutate(vocev, program = "Job Training"),
+  mutate(work, program = "Work"),
+  mutate(edev, program = "Education")
+)
+
+# Convert region to factor with explicit levels
+programs_data$region <- factor(programs_data$region, levels = 1:4, labels = c("Northeast", "Midwest", "South", "West"))
+
+# Create the figure using ggplot2
+ggplot(programs_data, aes(x = year, y = xbar * 100, group = region, color = region, shape = region)) +
+  geom_line(size = 1) +
+  geom_point(size = 3) +
[…]

### 2024-02-14 paper/paper.qmd
Add Aamishi as a contributor

@@
   - Kaavya Kalani
+  - Aamishi Avarsekar
 thanks: "Code and data are available at: LINK."

### 2024-02-14 paper/paper.qmd
Replication and github link added

--- 2024-02-13 paper/paper.qmd
@@
   - Kaavya Kalani
-thanks: "Code and data are available at: LINK."
+  - Aamishi Avarsekar
+thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
 date: today

### 2024-02-14 scripts/results.R
graphs for the results section

@@ -1,0 +1,162 @@
+install.packages("opendatatoronto")
+install.packages("knitr")
+install.packages("lubridate")
+install.packages("tidyverse")
+install.packages("janitor")
+install.packages("arrow")
+install.packages("gridExtra")
+
+library(arrow)
+library(knitr)
+library(janitor)
+library(lubridate)
+library(opendatatoronto)
+library(tidyverse)
+library(gridExtra)
+
+
+# general health and conditions in prisons
+health_data_count_by_year <- read_csv('data/msfp0116stt02data.csv')
+health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')
+
+health_demographic_data_clean <-
+  clean_names(health_data_count_by_demographics) |>
+  select("x1", "cancer", "heart_disease", "liver_disease", 
+         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
+         "drug_alcohol_intoxication", "accident", "homicide_d") 
+
+head(health_demographic_data_clean)
+
+health_demographic_data_long <- pivot_longer(health_demographic_data_clean, cols = -x1, names_to = "Cause", values_to = "Count")
+
+filtered_df <- health_demographic_data_clean[health_demographic_data_clean$x1 %in% c("Male", "Female"),]
+# all variables
+ggplot(health_demographic_data_long, aes(x = x1, y = Count, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Causes of Death by Category",
+       x = "Category",
+       y = "Count") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# genders: male and female
+gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))
+# transform
+gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")
+
+ggplot(gender_health_data_long, aes(x = x1, y = Count, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Causes of Death by Category",
+       x = "Category",
+       y = "Count") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggplot(gender_health_data_long, aes(x = x1, y = Count, fill = Cause)) +
+  geom_col(position = "dodge") +
+  facet_wrap(~x1, scales = "free_x") +
[…]

### 2024-02-14 scripts/graph1.R
Replication and github link added

@@ -1,0 +1,31 @@
+# install.packages("tidyverse")
+# install.packages("janitor")
+
+library(janitor)
+library(tidyverse)
+
+health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')
+
+health_demographic_data_clean <-
+  clean_names(health_data_count_by_demographics) |>
+    select("x1", "cancer", "heart_disease", "liver_disease",
+           "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
+           "drug_alcohol_intoxication", "accident", "homicide_d")
+
+gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))
+
+gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")
+
+# Calculate proportions within each category
+gender_health_data_long <- gender_health_data_long %>%
+  group_by(x1) %>%
+  mutate(Proportion = Count / sum(Count))
+
+# Plot the proportions graph using geom_bar with position = "fill"
+ggplot(gender_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(title = "Proportions of Causes of Death by Category",
+       x = "Gender",
+       y = "Proportion") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))

### 2024-02-14 scripts/99-replication.R
Update 99-replication.R

@@ -24,7 +24,7 @@ library(zoo)
 
 # Figure 2 replication
 
-x <- read_dta("../data/spi86to16.dta")
+x <- read_dta("data/spi86to16.dta")
 x <- x[x$admit == 1, ]
 
 # Create data frames for each program
@@ -64,7 +64,7 @@ State Prisoners, by Region") +
 pop <- c(1217352, 412783, 328239523)  # population denominators
 
 # Read CSV file with the correct date format
-x <- read.csv("../data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
+x <- read.csv("data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
 x$date <- as.Date(x$date, format = "%m/%d/%Y")  # Specify the correct date format
 
 # Calculate rolling averages
@@ -99,13 +99,13 @@ Those in Prison, Prison Staff, and in the General Population: May to September
 # Figure 4 replication
 
 # Read state covid data
-x <- read.csv("../data/all-states-history.csv", header = TRUE)
+x <- read.csv("data/all-states-history.csv", header = TRUE)
 
 # Data from 1/22/21
 x122 <- x[x$date == "2021-01-22",]
 
 # Read state population data
-pop <- read.csv("../data/statepop.csv", header = FALSE)
+pop <- read.csv("data/statepop.csv", header = FALSE)
 
 # Matching states
 ind <- match(x122$state, pop[,2])
@@ -115,7 +115,7 @@ x122$pop <- pop[ind,3]
 x122$ccr <- x122$positive / (x122$pop * 1e6)
 
 # Read prison covid data
-x1 <- read.csv("../data/COVID-19 Cases in State and Federal Prison Systems.csv", header = TRUE)
+x1 <- read.csv("data/COVID-19 Cases in State and Federal Prison Systems.csv", header = TRUE)
 
 # Data cleaning
 x1 <- lapply(x1, function(i) { sub(",", "", i) })
@@ -156,7 +156,3 @@ Population, by State: March 2020 to January 2021") +
   scale_x_continuous(limits = c(0, 800)) +
   scale_y_continuous(breaks = 1:nrow(ccr), labels = ccr$state, position = "right") +
   guides(color = guide_legend(title = NULL, override.aes = list(shape = c(16, 16))))
-
-
-
-

### 2024-02-15 paper/paper.qmd
2 graphs added and qmd structure improved

--- 2024-02-13 paper/paper.qmd
@@
 title: "Racial Force Probe"
-subtitle: "The study looks on a single scientific argument about whether or not certain police officers discriminate against people based on their race. "
 author:
@@
 
+#### Preamble ####
+# Purpose: Read the cleaned dataset to use it for the analysis
+# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
+# Date: 14 February 2024
+# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: run scripts/01-download_data.R followed by scripts/02-data_cleaning.R
+
+#### Workspace setup ####
+# install.packages("tidyverse")
+# install.packages("knitr")
+# install.packages("here")
+# install.packages("ggplot2")
+# install.packages("janitor")
+
+library(ggplot2)
 library(tidyverse)
-library(palmerpenguins)
+library(knitr)
+library(here)
+library(janitor)
 ```
@@
 # Introduction
+
 Health and Safety of Prisoners Today
@@
 
-# Data {#sec-data}
-
-# Model
+# Data
 
 # Results
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-overall-death
+#| fig-cap: Causes of deaths among the prisoners
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-covid
+#| fig-cap: Covid trends in the jail
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-death-vs-gender
+#| fig-cap: Causes of deaths among the prisoners by gender
+
+health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')
+
+health_demographic_data_clean <-
+  clean_names(health_data_count_by_demographics) |>
+  select("x1", "cancer", "heart_disease", "liver_disease",
+         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
+         "drug_alcohol_intoxication", "accident", "homicide_d")
+
+# Create a mapping for all causes
+cause_mapping <- c(
+  "cancer" = "Cancer",
+  "heart_disease" = "Heart Disease",
+  "liver_disease" = "Liver Disease",
+  "respiratory_disease" = "Respiratory Disease",
+  "aids_related" = "Aid related",
+  "all_other_illnesses" = "All other illnesses",
+  "suicide" = "Suicide",
+  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
+  "accident" = "Accident",
+  "homicide_d" = "Homicide"
+)
+
+# Rename the causes in the dataset
+health_demographic_data_clean <- health_demographic_data_clean %>%
+  rename_with(~ cause_mapping[.], -x1)
+
+gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))
+
+gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")
+
+# Calculate proportions within each category
+gender_health_data_long <- gender_health_data_long %>%
+  group_by(x1) %>%
+  mutate(Proportion = Count / sum(Count))
+
+# Plot the proportions graph using geom_bar with position = "fill"
+ggplot(gender_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
+  labs(title = "Proportions of Causes of Death by Category",
+       x = "Gender",
+       y = "Proportion") +
+  scale_fill_discrete(labels = cause_mapping) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-death-vs-race
+#| fig-cap: Causes of deaths among the prisoners by race
+
+# Read data
+health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')
+
+# Clean and organize data
+health_demographic_data_clean <-
+  clean_names(health_data_count_by_demographics) |>
+  select("x1", "cancer", "heart_disease", "liver_disease",
+         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
+         "drug_alcohol_intoxication", "accident", "homicide_d")
+
+# Pivot data to long format
+health_demographic_data_long <- pivot_longer(health_demographic_data_clean, cols = -x1, names_to = "Cause", values_to = "Count")
+
+# Filter data for specific demographics (e.g., "Male" and "Female")
+filtered_df <- health_demographic_data_clean[health_demographic_data_clean$x1 %in% c("Male", "Female"),]
+
+# Filter data for specific races (e.g., "White", "Black", "Hispanic", "Other")
+race_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("White", "Black", "Hispanic", "Other"))
+
+# Pivot data to long format for race
+race_health_data_long <- pivot_longer(race_health_data, cols = -x1, names_to = "Cause", values_to = "Count")
+
+# Calculate proportions instead of counts
+race_health_data_long <- race_health_data_long %>%
+  group_by(x1) %>%
+  mutate(Proportion = Count / sum(Count))
+
+# Create a mapping for all causes
+cause_mapping <- c(
+  "cancer" = "Cancer",
+  "heart_disease" = "Heart Disease",
+  "liver_disease" = "Liver Disease",
+  "respiratory_disease" = "Respiratory Disease",
+  "aids_related" = "Aid related",
+  "all_other_illnesses" = "All other illnesses",
+  "suicide" = "Suicide",
+  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
+  "accident" = "Accident",
+  "homicide_d" = "Homicide"
+)
+
+# Rename the causes in the dataset
+race_health_data_long <- race_health_data_long %>%
+  mutate(Cause = cause_mapping[as.character(Cause)])
+
+# Create a bar plot showing proportions with black borders
+ggplot(race_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
+  labs(title = "Proportions of Causes of Death by Category",
+       x = "Category",
+       y = "Proportion") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
 
@@
 
-# Appendix {-}
-
-# Additional data details
+# Appendix
 

### 2024-02-15 scripts/graph1.R

@@ -1,31 +1,0 @@
-# install.packages("tidyverse")
-# install.packages("janitor")
-
-library(janitor)
-library(tidyverse)
-
-health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')
-
-health_demographic_data_clean <-
-  clean_names(health_data_count_by_demographics) |>
-    select("x1", "cancer", "heart_disease", "liver_disease",
-           "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
-           "drug_alcohol_intoxication", "accident", "homicide_d")
-
-gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))
-
-gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")
-
-# Calculate proportions within each category
-gender_health_data_long <- gender_health_data_long %>%
-  group_by(x1) %>%
-  mutate(Proportion = Count / sum(Count))
-
-# Plot the proportions graph using geom_bar with position = "fill"
-ggplot(gender_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Proportions of Causes of Death by Category",
-       x = "Gender",
-       y = "Proportion") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))

### 2024-02-15 paper/paper.qmd
file paths updated

@@
 
-health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')
+health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))
 
@@
 # Read data
-health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')
+health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))
 

### 2024-02-15 scripts/99-replication.R

@@ -24,7 +24,7 @@ library(zoo)
 
 # Figure 2 replication
 
-x <- read_dta("data/spi86to16.dta")
+x <- read_dta("data/raw_data/spi86to16.dta")
 x <- x[x$admit == 1, ]
 
 # Create data frames for each program
@@ -64,7 +64,7 @@ State Prisoners, by Region") +
 pop <- c(1217352, 412783, 328239523)  # population denominators
 
 # Read CSV file with the correct date format
-x <- read.csv("data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
+x <- read.csv("data/raw_data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
 x$date <- as.Date(x$date, format = "%m/%d/%Y")  # Specify the correct date format
 
 # Calculate rolling averages
@@ -99,13 +99,13 @@ Those in Prison, Prison Staff, and in the General Population: May to September
 # Figure 4 replication
 
 # Read state covid data
-x <- read.csv("data/all-states-history.csv", header = TRUE)
+x <- read.csv("data/raw_data/all-states-history.csv", header = TRUE)
 
 # Data from 1/22/21
 x122 <- x[x$date == "2021-01-22",]
 
 # Read state population data
-pop <- read.csv("data/statepop.csv", header = FALSE)
+pop <- read.csv("data/raw_data/statepop.csv", header = FALSE)
 
 # Matching states
 ind <- match(x122$state, pop[,2])
@@ -115,7 +115,7 @@ x122$pop <- pop[ind,3]
 x122$ccr <- x122$positive / (x122$pop * 1e6)
 
 # Read prison covid data
-x1 <- read.csv("data/COVID-19 Cases in State and Federal Prison Systems.csv", header = TRUE)
+x1 <- read.csv("data/raw_data/COVID-19 Cases in State and Federal Prison Systems.csv", header = TRUE)
 
 # Data cleaning
 x1 <- lapply(x1, function(i) { sub(",", "", i) })

### 2024-02-15 paper/paper.qmd
new files

@@
 # License: MIT
-# Pre-requisites: run scripts/01-download_data.R followed by scripts/02-data_cleaning.R
+# Pre-requisites: run scripts/01-data_cleaning.R
 

### 2024-02-15 paper/paper.qmd
Graph added

@@
 # install.packages("janitor")
+# install.packages("zoo")
 
@@
 library(janitor)
+library(zoo)
 ```
@@
 #| fig-cap: Covid trends in the jail
+
+pop <- c(1217352, 412783, 328239523)  # population denominators
+
+# Read CSV file with the correct date format
+x <- read.csv("data/raw_data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
+x$date <- as.Date(x$date, format = "%m/%d/%Y")  # Specify the correct date format
+
+# Calculate rolling averages
+pp <- 1e5 * zoo::rollapply((x$pp_cases[-1] - x$pp_cases[-132]), width = 7, FUN = mean) / pop[1]
+sp <- 1e5 * zoo::rollapply((x$sp_cases[-1] - x$sp_cases[-132]), width = 7, FUN = mean) / pop[2]
+gp <- 1e5 * zoo::rollapply((x$gp_cases[-1] - x$gp_cases[-132]), width = 7, FUN = mean) / pop[3]
+
+# Combine data into a data frame
+ts_data <- data.frame(
+  date = rep(x$date[-(1:7)], 3),
+  new_cases_per_100k = c(pp, sp, gp),
+  group = rep(c("Incarcerated", "Prison Staff", "General Population"), each = length(pp))
+)
+
+# Create the figure using ggplot2 with solid lines
+ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
+  geom_line(size = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
+  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
+  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
+  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
+Those in Prison, Prison Staff, and in the General Population: May to September
+2020") +
+  theme_minimal() +
+  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
+             linetype = "dashed", color = "gray") +
+  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
+           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
+           vjust = -0.5, hjust = 0, color = "gray") +
+  scale_color_manual(values = c("blue", "red", "green"))
+
 ```

### 2024-02-15 scripts/results.R
general health & proportion

@@ -5,6 +5,7 @@ install.packages("tidyverse")
 install.packages("janitor")
 install.packages("arrow")
 install.packages("gridExtra")
+install.packages("readxl")
 
 library(arrow)
 library(knitr)
@@ -13,10 +14,108 @@ library(lubridate)
 library(opendatatoronto)
 library(tidyverse)
 library(gridExtra)
+library(readxl)
 
+cause_mapping <- c(
+  "cancer" = "Cancer",
+  "heart_disease" = "Heart Disease",
+  "liver_disease" = "Liver Disease",
+  "respiratory_disease" = "Respiratory Disease",
+  "aids_related" = "Aid related",
+  "all_other_illnesses" = "All other illnesses",
+  "suicide" = "Suicide",
+  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
+  "accident" = "Accident",
+  "homicide_d" = "Homicide"
+)
+
+# prisoner's proportion:
+prisoner_stats_data <- read_excel('data/mass_incarceration.xlsx')
+prisoner_necessary_year_data <-  prisoner_stats_data %>% filter(Year %in% c("2006", "2007", "2008","2009", "2010", 
+                                                                            "2011", "2012", "2013", "2014", "2015", "2016"))
+prisoner_stats_data_cleaned <- 
+  prisoner_necessary_year_data |>
+  select("Year", "Prison") |>
+  set_names("Year", "Prisoner_Count")
+
+tibble(prisoner_stats_data_cleaned)
 
 # general health and conditions in prisons
 health_data_count_by_year <- read_csv('data/msfp0116stt02data.csv')
+health_data_by_year_clean <-
+  clean_names(health_data_count_by_year) |>
+  select("x1", "x2006", "x2007", "x2008","x2009", "x2010", 
+         "x2011", "x2012", "x2013", "x2014", "x2015", "x2016") |>
+  set_names("Disease", "2006", "2007", "2008","2009", "2010", 
+         "2011", "2012", "2013", "2014", "2015", "2016")
+
+health_data_by_year <- health_data_by_year_clean %>% filter(Disease %in% c("Cancer", "Heart disease", "Liver disease",
+                                                                     "Respiratory disease", "AIDS-related", "All other illnessess", "Suicide",
+                                                                     "Drug/alcohol intoxication", "Accident", "Homicide", "Other causes"))
[…]

### 2024-02-15 scripts/results.R
diseases by proportion graph

@@ -55,49 +55,32 @@ health_data_by_year <- health_data_by_year_clean %>% filter(Disease %in% c("Canc
 # transform
 health_data_by_year_long <- pivot_longer(health_data_by_year, cols = -Disease, names_to = "Year", values_to = "Count")
 
+# this includes the prisoner count for each year
 merged_data <- merge(health_data_by_year_long, prisoner_stats_data_cleaned, by = "Year")
 
-
+# generating proportions: health_data_by_year_long by proportion
 health_data_by_year_long_prop <- merged_data %>%
-  group_by(Disease) %>%
-  mutate(Proportion = Count / sum(Count))
+  mutate(Proportion = Count / Prisoner_Count)
 
-ggplot(health_data_by_year_long_p, aes(x = Disease, y = Proportion, fill = Disease)) +
-  geom_bar(stat = "identity", position = "stack", size = 0.8) +  # Add black borders
+# can remove if we don't need this graph
+ggplot(health_data_by_year_long_prop, aes(x = Disease, y = Proportion, fill = Disease)) +
+  geom_bar(stat = "identity", position = "stack", size = 0.8) +  
   labs(title = "Proportions of Causes of Death by Category",
-       x = "Gender",
+       x = "Disease",
        y = "Proportion") +
   scale_fill_discrete(labels = cause_mapping) +
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
-data_long <- pivot_longer(health_data_by_year, cols = -Disease, names_to = "Year", values_to = "Cases")
 
-
-ggplot(data_long, aes(x = as.numeric(Year), y = Cases, color = Disease)) +
-  geom_line() +
+ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
+  geom_line(size = 1.0) +  
   labs(title = "Trends of Diseases by Year",
        x = "Year",
-       y = "Number of Cases",
+       y = "Proportion of cases",
        color = "Disease") +
   theme_minimal()
 
-ggplot(data_long, aes(x = as.numeric(Year), y = Cases, color = Disease)) +
-  geom_line(size = 1.5) +  # Adjust the size parameter for thicker lines
-  labs(title = "Trends of Diseases by Year",
-       x = "Year",
-       y = "Number of Cases",
-       color = "Disease") +
-  theme_minimal()
-
[…]

### 2024-02-15 paper/paper.qmd
Data cleaning done and graphs completed

@@
 # install.packages("ggplot2")
-# install.packages("janitor")
-# install.packages("zoo")
+# install.packages("readxl")
 
 library(ggplot2)
+library(readxl)
 library(tidyverse)
@@
 library(here)
-library(janitor)
-library(zoo)
 ```
@@
 
+## Source
+For this analysis we have used multiple datasets.
+This paper uses the datasets available in the replication package of the original paper.
+The author of the original paper has collated a large range of files from multiple sources and we will be using some of the datasets from this collation.
+
+## Methodology
+For this paper, the datasets were cleaned and analysed using the statistical programming software R [@citeR] along with the help tidyverse [@citeTidyverse], knitr [@citeKnitr], ggplot2 [@citeGgplot], here [@citeHere], zoo [@citeZoo], readxl [@citeRead], writexl [@citeWrite] and janitor [@citeJanitor].
+
+## Datasets
+### mass-incarceration.xlsx
+
+This dataset is a compilation of data collated by the original paper’s author from different sources:
+- Sourcebook of Criminal Justice Statistics
+- Bureau of Justice Statistics Report
+- Jail Inmates in 2018
+- U.S. Population FRED Data
+- World Prison Population Data
+- Western B. and Pettit B data
+
+There are three sheets in this file, we will be using sheet 1 which contains time series data for incarceration from 1925-2019; each year is associated with the number of imprisoned individuals per 100k people, calculated from the total population and total number of inmates.
+We will use this dataset to get the counts of prisons for different years. The Prison variable in the dataset is that variable.
+
+### msfp0116stt02data.csv and msfp0116stt09data.csv
+The original paper’s author got this data from
+- the BJS publication by E. Ann Carson, Mary P. Cowhig (2020), Mortality in State And Federal Prisons
+- 2001-2016 Statistical Tables (NCJ 251920)
+
+msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016.
+These causes include blah blah blah.
+For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over
+
+msfp0116stt09data.csv contains counts for the same conditions but segmented by demographic factors instead of time, such as gender, race, and age.
+For our analysis, we will be using the stats segmented by gender and race.
+
+### covidts01.csv
+The original paper’s author got this data from
+- Covid Prison Project (https://covidprisonproject.com) by Lauren BrinkelyRubinstein (UNC Chapel Hill) and Kathryn Nowotny (University of Miami).
+
+This dataset contains the collective number of new cases and case rates per 100,000 among the incarcerated, prison staff, and the general population.
+The numbers are recorded on a daily basis from April 22, 2020 to the end of August 2020.
+
+We will now analyse these datasets to uncover trends and draw conclusions.
+
 # Results
@@
 #| fig-cap: Causes of deaths among the prisoners
+
+# Load cleaned data from CSV
+health_data_by_year_long_prop <- read.csv("data/analysis_data/cleaned_health_data_by_year.csv", header = TRUE, stringsAsFactors = FALSE)
+
+# Create the plot
+ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
+  geom_line(size = 1.0) +
+  labs(title = "Trends of Diseases by Year",
+       x = "Year",
+       y = "Proportion of cases",
+       color = "Disease") +
+  theme_minimal()
+
 ```
@@
 
-pop <- c(1217352, 412783, 328239523)  # population denominators
-
-# Read CSV file with the correct date format
-x <- read.csv("data/raw_data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
-x$date <- as.Date(x$date, format = "%m/%d/%Y")  # Specify the correct date format
-
-# Calculate rolling averages
-pp <- 1e5 * zoo::rollapply((x$pp_cases[-1] - x$pp_cases[-132]), width = 7, FUN = mean) / pop[1]
-sp <- 1e5 * zoo::rollapply((x$sp_cases[-1] - x$sp_cases[-132]), width = 7, FUN = mean) / pop[2]
-gp <- 1e5 * zoo::rollapply((x$gp_cases[-1] - x$gp_cases[-132]), width = 7, FUN = mean) / pop[3]
-
-# Combine data into a data frame
-ts_data <- data.frame(
-  date = rep(x$date[-(1:7)], 3),
-  new_cases_per_100k = c(pp, sp, gp),
-  group = rep(c("Incarcerated", "Prison Staff", "General Population"), each = length(pp))
-)
+# Load cleaned data
+ts_data <- read.csv("data/analysis_data/cleaned_covid19_data.csv", header = TRUE, stringsAsFactors = FALSE)
 
@@
 
-health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))
+# Load cleaned data
+health_demographic_data_clean <- read_csv("data/analysis_data/cleaned_health_data_gender.csv")
 
-health_demographic_data_clean <-
-  clean_names(health_data_count_by_demographics) |>
-  select("x1", "cancer", "heart_disease", "liver_disease",
-         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
-         "drug_alcohol_intoxication", "accident", "homicide_d")
-
-# Create a mapping for all causes
-cause_mapping <- c(
-  "cancer" = "Cancer",
-  "heart_disease" = "Heart Disease",
-  "liver_disease" = "Liver Disease",
-  "respiratory_disease" = "Respiratory Disease",
-  "aids_related" = "Aid related",
-  "all_other_illnesses" = "All other illnesses",
-  "suicide" = "Suicide",
-  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
-  "accident" = "Accident",
-  "homicide_d" = "Homicide"
-)
-
-# Rename the causes in the dataset
-health_demographic_data_clean <- health_demographic_data_clean %>%
-  rename_with(~ cause_mapping[.], -x1)
-
-gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))
-
-gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")
-
-# Calculate proportions within each category
-gender_health_data_long <- gender_health_data_long %>%
-  group_by(x1) %>%
-  mutate(Proportion = Count / sum(Count))
-
-# Plot the proportions graph using geom_bar with position = "fill"
-ggplot(gender_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
-  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
+# Plot the proportions graph using geom_bar with position = "dodge"
+ggplot(health_demographic_data_clean, aes(x = x1, y = Proportion, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
   labs(title = "Proportions of Causes of Death by Category",
@@
 
-# Read data
-health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))
-
-# Clean and organize data
-health_demographic_data_clean <-
-  clean_names(health_data_count_by_demographics) |>
-  select("x1", "cancer", "heart_disease", "liver_disease",
-         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
-         "drug_alcohol_intoxication", "accident", "homicide_d")
-
-# Pivot data to long format
-health_demographic_data_long <- pivot_longer(health_demographic_data_clean, cols = -x1, names_to = "Cause", values_to = "Count")
-
-# Filter data for specific demographics (e.g., "Male" and "Female")
-filtered_df <- health_demographic_data_clean[health_demographic_data_clean$x1 %in% c("Male", "Female"),]
-
-# Filter data for specific races (e.g., "White", "Black", "Hispanic", "Other")
-race_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("White", "Black", "Hispanic", "Other"))
-
-# Pivot data to long format for race
-race_health_data_long <- pivot_longer(race_health_data, cols = -x1, names_to = "Cause", values_to = "Count")
-
-# Calculate proportions instead of counts
-race_health_data_long <- race_health_data_long %>%
-  group_by(x1) %>%
-  mutate(Proportion = Count / sum(Count))
-
-# Create a mapping for all causes
-cause_mapping <- c(
-  "cancer" = "Cancer",
-  "heart_disease" = "Heart Disease",
-  "liver_disease" = "Liver Disease",
-  "respiratory_disease" = "Respiratory Disease",
-  "aids_related" = "Aid related",
-  "all_other_illnesses" = "All other illnesses",
-  "suicide" = "Suicide",
-  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
-  "accident" = "Accident",
-  "homicide_d" = "Homicide"
-)
-
-# Rename the causes in the dataset
-race_health_data_long <- race_health_data_long %>%
-  mutate(Cause = cause_mapping[as.character(Cause)])
+# Load cleaned data from CSV
+race_health_data_long <- read.csv("data/analysis_data/cleaned_health_data_race.csv", header = TRUE, stringsAsFactors = FALSE)
 
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
 ```

### 2024-02-15 scripts/results.R
References updated

@@ -1,244 +1,0 @@
-install.packages("opendatatoronto")
-install.packages("knitr")
-install.packages("lubridate")
-install.packages("tidyverse")
-install.packages("janitor")
-install.packages("arrow")
-install.packages("gridExtra")
-install.packages("readxl")
-
-library(arrow)
-library(knitr)
-library(janitor)
-library(lubridate)
-library(opendatatoronto)
-library(tidyverse)
-library(gridExtra)
-library(readxl)
-
-cause_mapping <- c(
-  "cancer" = "Cancer",
-  "heart_disease" = "Heart Disease",
-  "liver_disease" = "Liver Disease",
-  "respiratory_disease" = "Respiratory Disease",
-  "aids_related" = "Aid related",
-  "all_other_illnesses" = "All other illnesses",
-  "suicide" = "Suicide",
-  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
-  "accident" = "Accident",
-  "homicide_d" = "Homicide"
-)
-
-# prisoner's proportion:
-prisoner_stats_data <- read_excel('data/mass_incarceration.xlsx')
-prisoner_necessary_year_data <-  prisoner_stats_data %>% filter(Year %in% c("2006", "2007", "2008","2009", "2010", 
-                                                                            "2011", "2012", "2013", "2014", "2015", "2016"))
-prisoner_stats_data_cleaned <- 
-  prisoner_necessary_year_data |>
-  select("Year", "Prison") |>
-  set_names("Year", "Prisoner_Count")
-
-tibble(prisoner_stats_data_cleaned)
-
-# general health and conditions in prisons
-health_data_count_by_year <- read_csv('data/msfp0116stt02data.csv')
-health_data_by_year_clean <-
-  clean_names(health_data_count_by_year) |>
-  select("x1", "x2006", "x2007", "x2008","x2009", "x2010", 
-         "x2011", "x2012", "x2013", "x2014", "x2015", "x2016") |>
-  set_names("Disease", "2006", "2007", "2008","2009", "2010", 
-         "2011", "2012", "2013", "2014", "2015", "2016")
-
-health_data_by_year <- health_data_by_year_clean %>% filter(Disease %in% c("Cancer", "Heart disease", "Liver disease",
-                                                                     "Respiratory disease", "AIDS-related", "All other illnessess", "Suicide",
[…]

### 2024-02-15 paper/paper.qmd
Data section added

@@
 The author of the original paper has collated a large range of files from multiple sources and we will be using some of the datasets from this collation.
+The raw datasets obtained from the replication package and cleaned and combined to form the datasets for this paper
 
@@
 ## Datasets
-### mass-incarceration.xlsx
 
-This dataset is a compilation of data collated by the original paper’s author from different sources:
-- Sourcebook of Criminal Justice Statistics
-- Bureau of Justice Statistics Report
-- Jail Inmates in 2018
-- U.S. Population FRED Data
-- World Prison Population Data
-- Western B. and Pettit B data
+### cleaned_health_data_by_year.csv
+This dataset is obtained by cleaning and combining the raw mass-incarceration.xlsx file and msfp0116stt02data.csv.
 
-There are three sheets in this file, we will be using sheet 1 which contains time series data for incarceration from 1925-2019; each year is associated with the number of imprisoned individuals per 100k people, calculated from the total population and total number of inmates.
-We will use this dataset to get the counts of prisons for different years. The Prison variable in the dataset is that variable.
-
-### msfp0116stt02data.csv and msfp0116stt09data.csv
-The original paper’s author got this data from
-- the BJS publication by E. Ann Carson, Mary P. Cowhig (2020), Mortality in State And Federal Prisons
-- 2001-2016 Statistical Tables (NCJ 251920)
+mass-incarceration.xlsx contains three sheets from which we will be using sheet 1 which contains time series data for incarceration from 1925-2019; each year is associated with the number of imprisoned individuals per 100k people, calculated from the total population and total number of inmates.
+We clean this file to get the counts of prisoners for 2006-2016.
 
 msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016.
-These causes include blah blah blah.
-For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over
+For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over.
 
-msfp0116stt09data.csv contains counts for the same conditions but segmented by demographic factors instead of time, such as gender, race, and age.
+Cleaning and combining these lead to our  cleaned_health_data_by_year.csv which contains:
+- Year (between 2006 to 2016 inclusive)
+- Disease (the cause of death of the prisoner)
+- Count (the number of prisoners who died due the corresponding cause in the corresponding year)
+- Prisoner_Count (the number of prisoners in that year)
+- Proportion (the proportion of prisoners who died due the corresponding cause in the corresponding year)
+
+A person becomes an entry in this dataset if they were a prisoner in any of the prisons in the United States of America between 2006 and 2016 inclusive and died in that year.
+The reason of the death is recorded among the following causes’categories: Illness, Cancer, Heart Diseases, Liver Diseases, Respiratory disease, AIDS-related, Suicide, Drug/alcohol intoxication, Accident, Homicide, All other reasons or illnesses.
+
+### cleaned_health_data_race.csv and cleaned_health_data_gender.csv
+This dataset is obtained by cleaning the raw msfp0116stt09data.csv file.
+
+msfp0116stt09data.csv contains counts for  causes of deaths of the prisoners for the years 2001, 2006-2016 but segmented by demographic factors instead of time, such as gender, race, and age.
 For our analysis, we will be using the stats segmented by gender and race.
 
-### covidts01.csv
-The original paper’s author got this data from
-- Covid Prison Project (https://covidprisonproject.com) by Lauren BrinkelyRubinstein (UNC Chapel Hill) and Kathryn Nowotny (University of Miami).
+The information segmented by race leads to cleaned_health_data_race.csv which contains:
+- x1 (The race of the prisoner)
+- Cause (the cause of death of the prisoner)
+- Count (the number of prisoners who died due the corresponding cause in the corresponding year)
+- Proportion (the proportion of prisoners who died due the corresponding cause in the corresponding year)
 
-This dataset contains the collective number of new cases and case rates per 100,000 among the incarcerated, prison staff, and the general population.
+A person becomes an entry in this dataset if they were a prisoner in any of the prisons in the United States of America in 2001 or between 2006 and 2016 inclusive and died in that year.
+The reason of the death is recorded among the following causes’categories: Illness, Cancer, Heart Diseases, Liver Diseases, Respiratory disease, AIDS-related, Suicide, Drug/alcohol intoxication, Accident, Homicide, All other reasons or illnesses.
+The race of the person is classified as Black, Hispanic, White or Other
+
+The information segmented by race leads to cleaned_health_data_gender.csv which contains:
+- x1 (The gender of the prisoner)
+- Cause (the cause of death of the prisoner)
+- Count (the number of prisoners who died due the corresponding cause in the corresponding year)
+- Proportion (the proportion of prisoners who died due the corresponding cause in the corresponding year)
+
+A person becomes an entry in this dataset if they were a prisoner in any of the prisons in the United States of America in 2001 or between 2006 and 2016 inclusive and died in that year.
+The reason of the death is recorded among the following causes’categories: Illness, Cancer, Heart Diseases, Liver Diseases, Respiratory disease, AIDS-related, Suicide, Drug/alcohol intoxication, Accident, Homicide, All other reasons or illnesses.
+The gender of the person is either Male or Female in this dataset.
+
+### cleaned_covid19_data.csv
+This dataset is obtained by cleaning the raw covidts01.csv file.
+
+covidts01.csv contains the collective number of new cases and case rates per 100,000 among the incarcerated, prison staff, and the general population.
 The numbers are recorded on a daily basis from April 22, 2020 to the end of August 2020.
+For our analysis, we just look at the number of new cases per 100,000.
+
+- Cleaning of this file leads to cleaned_covid19_data.csv which contains:
+- Date (The date of the new case)
+- New_cases_per_100k (The number of new cases recorded on the specific date per 100k)
+- Group (The group of people the case belonged to)
+
+A person becomes an entry in this dataset if they were incarcerated or prison staff in any of the prisons or the general population in the United States of America between April 22, 2020 to the end of August 2020 inclusive and contracted Covid-19.
 

### 2024-02-15 paper/paper.qmd
initial draft for results section

--- 2024-02-15 paper/paper.qmd
@@
 title: "Racial Force Probe"
+subtitle: "The study looks on a single scientific argument about whether or not certain police officers discriminate against people based on their race. "
 author:
@@
   - Aamishi Avarsekar
-thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
+thanks: "Code and data are available at: LINK."
 date: today
@@
 
+library(tidyverse)
+library(palmerpenguins)
 #### Preamble ####
@@
 # install.packages("readxl")
+# install.packages("gdata")
+# install.packages("dplyr")
+# install.packages("haven")
 
@@
 library(here)
+library(gdata)
+library(dplyr)
+library(haven)
 ```
@@
 # Introduction
-
 Health and Safety of Prisoners Today
@@
 
-# Data
+# Data {#sec-data}
 
-## Source
-For this analysis we have used multiple datasets.
-This paper uses the datasets available in the replication package of the original paper.
-The author of the original paper has collated a large range of files from multiple sources and we will be using some of the datasets from this collation.
-
-## Methodology
-For this paper, the datasets were cleaned and analysed using the statistical programming software R [@citeR] along with the help tidyverse [@citeTidyverse], knitr [@citeKnitr], ggplot2 [@citeGgplot], here [@citeHere], zoo [@citeZoo], readxl [@citeRead], writexl [@citeWrite] and janitor [@citeJanitor].
-
-## Datasets
-### mass-incarceration.xlsx
-
-This dataset is a compilation of data collated by the original paper’s author from different sources:
-- Sourcebook of Criminal Justice Statistics
-- Bureau of Justice Statistics Report
-- Jail Inmates in 2018
-- U.S. Population FRED Data
-- World Prison Population Data
-- Western B. and Pettit B data
-
-There are three sheets in this file, we will be using sheet 1 which contains time series data for incarceration from 1925-2019; each year is associated with the number of imprisoned individuals per 100k people, calculated from the total population and total number of inmates.
-We will use this dataset to get the counts of prisons for different years. The Prison variable in the dataset is that variable.
-
-### msfp0116stt02data.csv and msfp0116stt09data.csv
-The original paper’s author got this data from
-- the BJS publication by E. Ann Carson, Mary P. Cowhig (2020), Mortality in State And Federal Prisons
-- 2001-2016 Statistical Tables (NCJ 251920)
-
-msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016.
-These causes include blah blah blah.
-For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over
-
-msfp0116stt09data.csv contains counts for the same conditions but segmented by demographic factors instead of time, such as gender, race, and age.
-For our analysis, we will be using the stats segmented by gender and race.
-
-### covidts01.csv
-The original paper’s author got this data from
-- Covid Prison Project (https://covidprisonproject.com) by Lauren BrinkelyRubinstein (UNC Chapel Hill) and Kathryn Nowotny (University of Miami).
-
-This dataset contains the collective number of new cases and case rates per 100,000 among the incarcerated, prison staff, and the general population.
-The numbers are recorded on a daily basis from April 22, 2020 to the end of August 2020.
-
-We will now analyse these datasets to uncover trends and draw conclusions.
+# Model
 
 # Results
+
+Based on the original paper’s findings, we decided to further investigate the health and general well-being of prisoners and trends of diseases within prison, causes of death and also the trends of COVID-19 across a 7-day period.
 
@@
 # Load cleaned data from CSV
-health_data_by_year_long_prop <- read.csv("data/analysis_data/cleaned_health_data_by_year.csv", header = TRUE, stringsAsFactors = FALSE)
+health_data_by_year_long_prop <- read.csv(file=here("data/analysis_data/cleaned_health_data_by_year.csv"), header = TRUE, stringsAsFactors = FALSE)
 
@@
 ```
+Figure 1 compares shows the cause of death among prisoners due to diseases.
+In order to accurately investigate, we adjusted the number of deaths caused by a particular disease by the total number of prisoners recorded for that year.
+This accounts for the differences in the number of deaths recorded between each cause.
+
+By observing the trends in Figure 1, we can see that the deaths due to cancer and heart diseases is significantly higher compared with other causes.
+The trends for death due to homicide saw a slight decrease between the later years which is opposite to the rate of deaths caused by suicide.
+While deaths caused by respiratory diseases remain fairly consistent, there is a gradual increase throughout the years.
+We can also observe that there is a steady decline in the deaths caused by AIDS.
+Deaths caused by accident remain fairly uncommon with only a very slight increase in 2012.
+However, the deaths caused by heart disease remained fairly high yet consistent but saw a steep incline after 2012.
+The deaths caused by cancer has the highest rate.
+To take a closer look at this, we can investigate the age-group of the prisoners.
 
@@
 # Load cleaned data
-ts_data <- read.csv("data/analysis_data/cleaned_covid19_data.csv", header = TRUE, stringsAsFactors = FALSE)
+ts_data <- read.csv(file=here("data/analysis_data/cleaned_covid19_data.csv"), header = TRUE, stringsAsFactors = FALSE)
+# Load cleaned data
+ts_data <- read.csv(file=here("data/analysis_data/cleaned_covid19_data.csv"), header = TRUE, stringsAsFactors = FALSE)
+
+#Convert date column to Date type
+ts_data$date <- as.Date(ts_data$date, format = "%Y-%m-%d")
 
@@
   labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
-Those in Prison, Prison Staff, and in the General Population: May to September
-2020") +
+#Those in Prison, Prison Staff, and in the General Population: May to September
+#2020") +
   theme_minimal() +
@@
 ```
+result 2
 
@@
 
+cause_mapping <- c(
+  "cancer" = "Cancer",
+  "heart_disease" = "Heart Disease",
+  "liver_disease" = "Liver Disease",
+  "respiratory_disease" = "Respiratory Disease",
+  "aids_related" = "Aid related",
+  "all_other_illnesses" = "All other illnesses",
+  "suicide" = "Suicide",
+  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
+  "accident" = "Accident",
+  "homicide_d" = "Homicide"
+)
+
 # Load cleaned data
-health_demographic_data_clean <- read_csv("data/analysis_data/cleaned_health_data_gender.csv")
+health_demographic_data_clean <- read_csv(file=here("data/analysis_data/cleaned_health_data_gender.csv"))
 
@@
 ```
+Figure 3 compares the cause of death due to diseases between female and male prisoners.
+The data in the graph has also been adjusted by proportion of female and male prisoners to account for the low counts of reported diseases in females as well as their population.
+Both graphs have a similar general unimodal shape.
+Both sexes have the highest deaths caused by cancer and heart diseases, with males with a slightly higher rate than females.
+However, the least deaths in females are caused by homicide where its rate is almost negligible.
+While homicide in males is significantly less compared to other death causes, it is still significantly higher compared to its female counter part.
+On the other hand, the least deaths in males occurs due to accident.
+Liver disease and respiratory disease are almost comparable in females whereas for males, there is a steep decline in deaths caused by respiratory diseases in comparison to deaths caused by liver disease.
 
@@
 # Load cleaned data from CSV
-race_health_data_long <- read.csv("data/analysis_data/cleaned_health_data_race.csv", header = TRUE, stringsAsFactors = FALSE)
+race_health_data_long <- read.csv(file=here("data/analysis_data/cleaned_health_data_race.csv"), header = TRUE, stringsAsFactors = FALSE)
 
@@
 
-# Discussion
+Figure 4 compares the causes of deaths in prison with respect to the different races of the prisoners.
+It can be observed that the cause of deaths are unimodular across all races but they have different peaks in their modes.
+Amongst Black prisoners, heart diseases have the highest death rate while in Hispanic, White and other races, cancer has the highest death rate.
+Amongst the identified races, Hispanic prisoners have the highest suicide and homicide rates and also respiratory diseases.
+It can also be observed that Black prisoners have the highest rate of aids-related deaths, and White prisoners have the highest rate of deaths caused by heart diseases.
+While deaths caused by drug/alcohol intoxication remains comparatively low across all races, it is the highest for Hispanic prisoners.
+Amongst other races, it can be observed that suicide is the highest recorded reason for death which could open way to an analyses for racism within prisons as well.
+
+## Discussion
 
@@
 
-# Appendix
+# Appendix {-}
+
+# Additional data details
 

### 2024-02-15 paper/paper-3.qmd
Add files via upload

--- 2024-02-15 paper/paper.qmd
@@
 ---
-title: "Racial Force Probe"
+title: "Prison Health"
 author:
@@
 date: today
+header-includes:
+    - \usepackage{setspace}\doublespacing
 date-format: long
@@
 The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being.
-Utilizing data collected over the past [insert number of years], this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners.
+Utilizing data collected over the past 15 years, this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners.
 By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being.
@@
 library(tidyverse)
+library(palmerpenguins)
+#### Preamble ####
+# Purpose: Read the cleaned dataset to use it for the analysis
+# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
+# Date: 14 February 2024
+# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: run scripts/01-data_cleaning.R
+
+#### Workspace setup ####
+# install.packages("tidyverse")
+# install.packages("knitr")
+# install.packages("here")
+# install.packages("ggplot2")
+# install.packages("readxl")
+# install.packages("gdata")
+# install.packages("dplyr")
+# install.packages("haven")
+
+library(ggplot2)
+library(readxl)
+library(tidyverse)
+library(knitr)
+library(here)
+library(gdata)
+library(dplyr)
+library(haven)
 library(knitr)
@@
 msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016.
-For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over.
+For our analysis we will not be considering 2001, but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over.
 
@@
 # Results
+Based on the original paper’s findings, we decided to further investigate the health and general well-being of prisoners and trends of diseases within prison, causes of death and also the trends of COVID-19 across a 7-day period.
 
@@
 # Load cleaned data from CSV
-health_data_by_year_long_prop <- read.csv("data/analysis_data/cleaned_health_data_by_year.csv", header = TRUE, stringsAsFactors = FALSE)
+
+health_data_by_year_long_prop <- read.csv(file=here("data/analysis_data/cleaned_health_data_by_year.csv"), header = TRUE, stringsAsFactors = FALSE)
 
@@
 ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
-  geom_line(size = 1.0) +
+  geom_line(linewidth = 1.0) +
   labs(title = "Trends of Diseases by Year",
@@
   theme_minimal()
+```
 
-```
+#Figure 1 
+compares shows the cause of death among prisoners due to diseases.
+In order to accurately investigate, we adjusted the number of deaths caused by a particular disease by the total number of prisoners recorded for that year.
+This accounts for the differences in the number of deaths recorded between each cause.
+
+By observing the trends in Figure 1, we can see that the deaths due to cancer and heart diseases is significantly higher compared with other causes.
+The trends for death due to homicide saw a slight decrease between the later years which is opposite to the rate of deaths caused by suicide.
+While deaths caused by respiratory diseases remain fairly consistent, there is a gradual increase throughout the years.
+We can also observe that there is a steady decline in the deaths caused by AIDS.
+Deaths caused by accident remain fairly uncommon with only a very slight increase in 2012.
+However, the deaths caused by heart disease remained fairly high yet consistent but saw a steep incline after 2012.
+The deaths caused by cancer has the highest rate.
+To take a closer look at this, we can investigate the age-group of the prisoners.
 
@@
 ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
-  geom_line(size = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
+  geom_line(linewidth = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
   scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
   scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
-  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
-Those in Prison, Prison Staff, and in the General Population: May to September
-2020") +
+  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases") +
   theme_minimal() +
@@
 
+#figure 2
+
 ```{r}
@@
 #| echo: false
-#| label: fig-death-vs-race
+#| label: fig-death-vs-race-2
 #| fig-cap: Causes of deaths among the prisoners by race
@@
 
+Figure 4 compares the causes of deaths in prison with respect to the different races of the prisoners.
+It can be observed that the cause of deaths are unimodular across all races but they have different peaks in their modes.
+Amongst Black prisoners, heart diseases have the highest death rate while in Hispanic, White and other races, cancer has the highest death rate.
+Amongst the identified races, Hispanic prisoners have the highest suicide and homicide rates and also respiratory diseases.
+It can also be observed that Black prisoners have the highest rate of aids-related deaths, and White prisoners have the highest rate of deaths caused by heart diseases.
+While deaths caused by drug/alcohol intoxication remains comparatively low across all races, it is the highest for Hispanic prisoners.
+Amongst other races, it can be observed that suicide is the highest recorded reason for death which could open way to an analyses for racism within prisons as well.
+
+## Discussion
+
+```
+
 # Discussion

### 2024-02-15 paper/paper-draft.qmd

--- 2024-02-15 paper/paper.qmd
@@
 
+Based on the original paper’s findings, we decided to further investigate the health and general well-being of prisoners and trends of diseases within prison, causes of death and also the trends of COVID-19 across a 7-day period.
+
 ```{r}
@@
   theme_minimal()
+
+```
+Figure 1 compares shows the cause of death among prisoners due to diseases.
+In order to accurately investigate, we adjusted the number of deaths caused by a particular disease by the total number of prisoners recorded for that year.
+This accounts for the differences in the number of deaths recorded between each cause.
+
+By observing the trends in Figure 1, we can see that the deaths due to cancer and heart diseases is significantly higher compared with other causes.
+The trends for death due to homicide saw a slight decrease between the later years which is opposite to the rate of deaths caused by suicide.
+While deaths caused by respiratory diseases remain fairly consistent, there is a gradual increase throughout the years.
+We can also observe that there is a steady decline in the deaths caused by AIDS.
+Deaths caused by accident remain fairly uncommon with only a very slight increase in 2012.
+However, the deaths caused by heart disease remained fairly high yet consistent but saw a steep incline after 2012.
+The deaths caused by cancer has the highest rate.
+To take a closer look at this, we can investigate the age-group of the prisoners.
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-covid
+#| fig-cap: Covid trends in the jail
+
+# Load cleaned data
+ts_data <- read.csv(file=here("data/analysis_data/cleaned_covid19_data.csv"), header = TRUE, stringsAsFactors = FALSE)
+# Load cleaned data
+ts_data <- read.csv(file=here("data/analysis_data/cleaned_covid19_data.csv"), header = TRUE, stringsAsFactors = FALSE)
+
+#Convert date column to Date type
+ts_data$date <- as.Date(ts_data$date, format = "%Y-%m-%d")
+
+# Create the figure using ggplot2 with solid lines
+ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
+  geom_line(size = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
+  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
+  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
+  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
+#Those in Prison, Prison Staff, and in the General Population: May to September
+#2020") +
+  theme_minimal() +
+  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
+             linetype = "dashed", color = "gray") +
+  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
+           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
+          vjust = -0.5, hjust = 0, color = "gray") +
+  scale_color_manual(values = c("blue", "red", "green"))
+
+```
+result 2
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-death-vs-gender
+#| fig-cap: Causes of deaths among the prisoners by gender
+
+cause_mapping <- c(
+  "cancer" = "Cancer",
+  "heart_disease" = "Heart Disease",
+  "liver_disease" = "Liver Disease",
+  "respiratory_disease" = "Respiratory Disease",
+  "aids_related" = "Aid related",
+  "all_other_illnesses" = "All other illnesses",
+  "suicide" = "Suicide",
+  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
+  "accident" = "Accident",
+  "homicide_d" = "Homicide"
+)
+
+# Load cleaned data
+health_demographic_data_clean <- read_csv(file=here("data/analysis_data/cleaned_health_data_gender.csv"))
+
+# Plot the proportions graph using geom_bar with position = "dodge"
+ggplot(health_demographic_data_clean, aes(x = x1, y = Proportion, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
+  labs(title = "Proportions of Causes of Death by Category",
+       x = "Gender",
+       y = "Proportion") +
+  scale_fill_discrete(labels = cause_mapping) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+Figure 3 compares the cause of death due to diseases between female and male prisoners.
+The data in the graph has also been adjusted by proportion of female and male prisoners to account for the low counts of reported diseases in females as well as their population.
+Both graphs have a similar general unimodal shape.
+Both sexes have the highest deaths caused by cancer and heart diseases, with males with a slightly higher rate than females.
+However, the least deaths in females are caused by homicide where its rate is almost negligible.
+While homicide in males is significantly less compared to other death causes, it is still significantly higher compared to its female counter part.
+On the other hand, the least deaths in males occurs due to accident.
+Liver disease and respiratory disease are almost comparable in females whereas for males, there is a steep decline in deaths caused by respiratory diseases in comparison to deaths caused by liver disease.
+
+```{r}
+#| message: false
+#| echo: false
+#| label: fig-death-vs-race
+#| fig-cap: Causes of deaths among the prisoners by race
+
+# Load cleaned data from CSV
+race_health_data_long <- read.csv(file=here("data/analysis_data/cleaned_health_data_race.csv"), header = TRUE, stringsAsFactors = FALSE)
+
+# Create a bar plot showing proportions with black borders
+ggplot(race_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
+  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
+  labs(title = "Proportions of Causes of Death by Category",
+       x = "Category",
+       y = "Proportion") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+Figure 4 compares the causes of deaths in prison with respect to the different races of the prisoners.
+It can be observed that the cause of deaths are unimodular across all races but they have different peaks in their modes.
+Amongst Black prisoners, heart diseases have the highest death rate while in Hispanic, White and other races, cancer has the highest death rate.
+Amongst the identified races, Hispanic prisoners have the highest suicide and homicide rates and also respiratory diseases.
+It can also be observed that Black prisoners have the highest rate of aids-related deaths, and White prisoners have the highest rate of deaths caused by heart diseases.
+While deaths caused by drug/alcohol intoxication remains comparatively low across all races, it is the highest for Hispanic prisoners.
+Amongst other races, it can be observed that suicide is the highest recorded reason for death which could open way to an analyses for racism within prisons as well.
+
+## Discussion
 

### 2024-02-15 scripts/00-simulate_data.R
simulation and basic tests

@@ -1,0 +1,115 @@
+#### Preamble ####
+# Purpose: Simulating the main datasets that will be used for this paper
+# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
+# Date: 15 February 2024
+# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+
+# Parts of gender simulation dataset been generated using ChatGPT3.5 and the chat
+# is included in the LLM usage
+
+# Set seed for reproducibility
+set.seed(123)
+
+# Generate random data for Male
+male_data <- data.frame(
+  Gender = "Male",
+  Cancer = rpois(1, lambda = 14000),
+  `Heart_Disease` = rpois(1, lambda = 13200),
+  `Liver_Disease` = rpois(1, lambda = 4800),
+  `Respiratory_Disease` = rpois(1, lambda = 3100),
+  `Aid_related` = rpois(1, lambda = 1800)
+)
+
+# Generate random data for Female
+female_data <- data.frame(
+  Gender = "Female",
+  Cancer = rpois(1, lambda = 550),
+  `Heart_Disease` = rpois(1, lambda = 460),
+  `Liver_Disease` = rpois(1, lambda = 170),
+  `Respiratory_Disease` = rpois(1, lambda = 170),
+  `Aid_related` = rpois(1, lambda = 100)
+)
+
+# Create data frame
+sim_data <- rbind(male_data, female_data)
+
+
+# Tests written by author (no LLM used):
+# None of the values must be null or missing
+
+# Gender:
+gender_test_1 <- all(sim_data$Cancer > 0) & all(sim_data$Heart_Disease > 0) &
+  all(sim_data$Liver_Disease > 0) & all(sim_data$Respiratory_Disease > 0
+                                        & all(sim_data$Aid_related > 0))
+stopifnot(gender_test_1 == TRUE)
+
+# Test data gender types
+stopifnot(is.numeric(sim_data$Cancer))
+stopifnot(is.numeric(sim_data$Heart_Disease))
+stopifnot(is.numeric(sim_data$Liver_Disease))
+stopifnot(is.numeric(sim_data$Respiratory_Disease))
+stopifnot(is.numeric(sim_data$Aid_related))
+
+
[…]

### 2024-02-16 paper/paper.qmd
removed double spacing

@@
 ---
-title: "Prison Health"
+title: "Behind Bars: Statistical Insights into the Wellbeing of U.S. Prisoners"
 author:

### 2024-02-16 paper/paper.qmd
Results completed

@@
 
-#### Preamble ####
-# Purpose: Read the cleaned dataset to use it for the analysis
-# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
-# Date: 14 February 2024
-# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: run scripts/01-data_cleaning.R
-
-#### Workspace setup ####
-# install.packages("tidyverse")
-# install.packages("knitr")
-# install.packages("here")
-# install.packages("ggplot2")
-# install.packages("readxl")
-
-library(ggplot2)
-library(readxl)
-library(tidyverse)
-library(palmerpenguins)
 #### Preamble ####
@@
 The author of the original paper has collated a large range of files from multiple sources and we will be using some of the datasets from this collation.
-The raw datasets obtained from the replication package and cleaned and combined to form the datasets for this paper
+The raw datasets obtained from the replication package and cleaned and combined to form the datasets for this paper.
 
@@
 Cleaning and combining these lead to our  cleaned_health_data_by_year.csv which contains:
+
 - Year (between 2006 to 2016 inclusive)
@@
 The information segmented by race leads to cleaned_health_data_race.csv which contains:
+
 - x1 (The race of the prisoner)
@@
 The information segmented by race leads to cleaned_health_data_gender.csv which contains:
+
 - x1 (The gender of the prisoner)
@@
 
-- Cleaning of this file leads to cleaned_covid19_data.csv which contains:
+Cleaning of this file leads to cleaned_covid19_data.csv which contains:
+
 - Date (The date of the new case)
@@
 
+\newpage
+
 # Results
-Based on the original paper’s findings, we decided to further investigate the health and general well-being of prisoners and trends of diseases within prison, causes of death and also the trends of COVID-19 across a 7-day period.
 
@@
 #| echo: false
+#| warning: false
 #| label: fig-overall-death
-#| fig-cap: Causes of deaths among the prisoners
+#| fig-cap: Causes of deaths among the prisoners from 2006 to 2016
 
@@
   geom_line(linewidth = 1.0) +
-  labs(title = "Trends of Diseases by Year",
-       x = "Year",
+  labs(x = "Year",
        y = "Proportion of cases",
@@
 
-#Figure 1 
-compares shows the cause of death among prisoners due to diseases.
-In order to accurately investigate, we adjusted the number of deaths caused by a particular disease by the total number of prisoners recorded for that year.
-This accounts for the differences in the number of deaths recorded between each cause.
-
-By observing the trends in Figure 1, we can see that the deaths due to cancer and heart diseases is significantly higher compared with other causes.
-The trends for death due to homicide saw a slight decrease between the later years which is opposite to the rate of deaths caused by suicide.
-While deaths caused by respiratory diseases remain fairly consistent, there is a gradual increase throughout the years.
+@fig-overall-death compares shows the cause of death among prisoners due to different causes from 2016 to 2019.
+We can see that the deaths due to cancer and heart diseases have significantly increased over the years and remain to cause significantly more deaths than any of the other causes.
 We can also observe that there is a steady decline in the deaths caused by AIDS.
+The trends for death due to homicide and drug/alcohol usage saw a slight increase over the years.
 Deaths caused by accident remain fairly uncommon with only a very slight increase in 2012.
-However, the deaths caused by heart disease remained fairly high yet consistent but saw a steep incline after 2012.
-The deaths caused by cancer has the highest rate.
-To take a closer look at this, we can investigate the age-group of the prisoners.
+The other causes have been almost the same with a few spikes over the years.
 
@@
 #| echo: false
+#| warning: false
 #| label: fig-covid
-#| fig-cap: Covid trends in the jail
+#| fig-cap: Moving Average of New Daily COVID-19 Cases in 2020
 
@@
   scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
-  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases") +
+  labs(x = "Month in 2020", y = "New cases per 100,000") +
   theme_minimal() +
@@
 
-#figure 2
+@fig-covid has been recreated from the original paper.
+This figure shows the number of new COVID-19 cases per 100,000 from April 22, 2020 to the end of August 2020.
+We notice that the number of new cases among the incarcerated is significantly more than the number of cases in the prison staff which is significantly more than the ones among the general population.
+We also see the general population population trend seems smoother compared to trend for the prison staff and the incarcerated.
 
@@
 #| echo: false
+#| warning: false
 #| label: fig-death-vs-gender
 #| fig-cap: Causes of deaths among the prisoners by gender
-
-cause_mapping <- c(
-  "cancer" = "Cancer",
-  "heart_disease" = "Heart Disease",
-  "liver_disease" = "Liver Disease",
-  "respiratory_disease" = "Respiratory Disease",
-  "aids_related" = "Aid related",
-  "all_other_illnesses" = "All other illnesses",
-  "suicide" = "Suicide",
-  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
-  "accident" = "Accident",
-  "homicide_d" = "Homicide"
-)
 
@@
   geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
-  labs(title = "Proportions of Causes of Death by Category",
-       x = "Gender",
+  labs(x = "Gender",
        y = "Proportion") +
-  scale_fill_discrete(labels = cause_mapping) +
   theme_minimal() +
@@
 ```
-Figure 3 compares the cause of death due to diseases between female and male prisoners.
-The data in the graph has also been adjusted by proportion of female and male prisoners to account for the low counts of reported diseases in females as well as their population.
-Both graphs have a similar general unimodal shape.
+
+@fig-death-vs-gender compares the cause of death due to diseases between female and male prisoners.
+Both graphs have a similar general shape.
 Both sexes have the highest deaths caused by cancer and heart diseases, with males with a slightly higher rate than females.
@@
 While homicide in males is significantly less compared to other death causes, it is still significantly higher compared to its female counter part.
+The proportion of females committing suicide in prison is also higher than males committing suicide.
 On the other hand, the least deaths in males occurs due to accident.
-Liver disease and respiratory disease are almost comparable in females whereas for males, there is a steep decline in deaths caused by respiratory diseases in comparison to deaths caused by liver disease.
+Liver disease and respiratory disease are almost comparable in females whereas for males, deaths caused by respiratory diseases are much lesser than deaths caused by liver disease.
 
@@
 #| echo: false
-#| label: fig-death-vs-race-2
+#| warning: false
+#| label: fig-death-vs-race
 #| fig-cap: Causes of deaths among the prisoners by race
@@
 ggplot(race_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
-  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
-  labs(title = "Proportions of Causes of Death by Category",
-       x = "Category",
+  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
+  labs(x = "Race",
        y = "Proportion") +
@@
 
-Figure 4 compares the causes of deaths in prison with respect to the different races of the prisoners.
-It can be observed that the cause of deaths are unimodular across all races but they have different peaks in their modes.
-Amongst Black prisoners, heart diseases have the highest death rate while in Hispanic, White and other races, cancer has the highest death rate.
-Amongst the identified races, Hispanic prisoners have the highest suicide and homicide rates and also respiratory diseases.
-It can also be observed that Black prisoners have the highest rate of aids-related deaths, and White prisoners have the highest rate of deaths caused by heart diseases.
-While deaths caused by drug/alcohol intoxication remains comparatively low across all races, it is the highest for Hispanic prisoners.
-Amongst other races, it can be observed that suicide is the highest recorded reason for death which could open way to an analyses for racism within prisons as well.
-
-## Discussion
-
-```
+@fig-death-vs-race compares the causes of deaths in prison with respect to the different races of the prisoners.
+Among Black prisoners, heart diseases have the highest death rate while in Hispanic, White and other races, cancer has the highest death rate.
+Among the identified races, Hispanic prisoners have the highest suicide and homicide rates and also respiratory diseases.
+It can also be observed that Black prisoners have the highest rate of aids-related deaths, much higher than any of the other races.
+Another interesting trend that can be observed is that suicide is much higher as the reason for death for other races than Black, Hispanic and White prisoners.
 
@@
 
-# Appendix
-
 # References

### 2024-02-16 paper/paper.qmd
added the updated abstract

@@
 date-format: long
-abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations.
-The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being.
-Utilizing data collected over the past 15 years, this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners.
-By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being.
-This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities.
-Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
+abstract: "This research explores the causes, consequences, and demographics of prisoner deaths in the United States.
+It reveals an increasing trend in the death of prisoners, highlighting heart problems and cancer as major causes.
+The frequency of cancer is shown by demographic study, with Black, Hispanic, and White inmates being most affected.
+The significant impact that prison staff has on the wellbeing of inmates is highlighted by gender-based differences
+in the availability of medical treatment and institutional discrimination.
+
+The study acknowledges the shortcomings in demographic representation and argues for a larger dataset.
+The study intends to shed light on aspects such as racism and hate crimes in correctional settings, underscoring the
+need for inclusive statistics on racial and sexual minorities. The request for more datasets is meant to strengthen the
+ research and provide a more comprehensive comprehension of the complex relationship between conditions in jail and overall health.
+
+In addition, the report recommends cross-national comparisons and an analysis of the criminal justice systems in
+countries with similar populations in order to evaluate general well-being and pandemic readiness. The study provides
+insight into the difficulties faced by prisoners and establishes a foundation for further investigations into systemic
+problems and healthcare disparities in correctional environments."
 format: pdf

### 2024-02-16 paper/paper.qmd
Abstract added, title updated

--- 2024-02-16 paper/paper.qmd
@@
 ---
-title: "Behind Bars: Statistical Insights into the Wellbeing of U.S. Prisoners"
+title: "Behind Bars: Insights into the Wellbeing of U.S. Prisoners"
 author:
@@
 date-format: long
-abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations.
-The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being.
-Utilizing data collected over the past 15 years, this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners.
-By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being.
-This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities.
-Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
+abstract: "This research explores the causes and trends of prisoner deaths in the United States giving insights into their wellbeing in the prisons.
+We examine these rates both from a comprehensive standpoint and delve into demographic breakdowns, specifically considering gender and race.
+The paper analyse trends revealing a notable rise in deaths due to cancer and heart diseases.
+While AIDS-related deaths decline, homicides and drug/alcohol-related deaths show a slight increase.
+Comparisons across gender highlight differences in causes of death, with males experiencing higher rates of homicide, while females show a greater propensity for suicide.
+The examination of racial disparities exposes varying leading causes of death, emphasizing the urgent need for targeted interventions to improve the overall wellbeing of incarcerated individuals.
+The trends uncovered in this paper establish a foundation for further investigations into systemic problems and healthcare disparities in correctional environments."
 format: pdf

### 2024-02-16 paper/paper.qmd
removed the abstract draft

@@
 # Introduction
-
-Health and Safety of Prisoners Today
-
-Individuals' health and safety in jail are critical aspects of the criminal justice system
-that must be thoroughly examined. As we negotiate the complicated world of prisons in contemporary America,
-it becomes clear that the physical and emotional health of jailed people is a major problem. This study aims to give a
-complete analysis of the health difficulties encountered by inmates today, diving into the areas of physical health,
-emotional well-being, death rates, and the subtle distinctions between male and female prisoners.
-
-Physical Health of Prisoners: Unveiling a Disturbing Reality
-
-The first aspect of our investigation is a thorough examination of the physical health of convicts.
-Recent discoveries in America portray a troubling picture, demonstrating that a sizable proportion of offenders enter
-correctional facilities with a variety of pre-existing physical health issues.
-The jail environment, which is marked by overcrowding, restricted access to healthcare, and a lack of preventive
-measures, frequently exacerbates these problems(Stevenson, B. (2021)). According to studies, the prison population has a high prevalence of
-chronic diseases such as diabetes, hypertension, and infectious diseases, emphasizing the critical need to focus on
-the physical well-being of jailed people.
-
-The sad fact is that convicts have far lower health outcomes than the general population.
-Access to medical care in prisons remains a recurrent issue, resulting in delayed diagnoses and insufficient treatment.
-The cumulative effect of these conditions adds to a cycle of declining physical health among inmates, raising not just
-ethical problems but also public health dangers when they are eventually released into society(M. Maruschak, L., & Berzofsky, M. (2015)).
-
-Mental Health: The Silent Crisis Behind Bars
-
-Moving beyond physical health, the second half of our investigation focuses on the sometimes-ignored topic of mental
-well-being in the prison system. Mental health difficulties among inmates have reached crisis proportions, with a slew
-of obstacles compounding the psychological toll of incarceration.(Edgemon, T. G., & Clay-Warner, J. (2019)). Suicide rates among convicts have become a major
-problem, underlining the critical need for mental health services in correctional facilities.
-
-Substance addiction exacerbates the mental health problem since many inmates take drugs to cope with the stresses of
-confinement. The absence of comprehensive addiction treatment services in jails reinforces the cycle of substance misuse,
-exacerbating mental health issues and impeding successful recovery attempts.(Edgemon, T. G., & Clay-Warner, J. (2019))
-
-The mental health crisis among inmates emphasizes the significance of treating the underlying causes of psychological
-anguish in correctional facilities. Initiatives to provide mental health support, provide counselling services, and
-establish drug addiction treatment programs are critical to interrupting the cycle of mental health worsening among inmates.
-
-From Suicides to Mortality Rates: Connecting the Dots
-
-The third feature of our investigation focuses on the relationship between suicides and prison death rates.
-Suicides in correctional facilities are not isolated instances; they are inextricably linked to larger concerns such as
-mental health, poor healthcare, and the harsh reality of incarceration. Beyond suicides, inmates are always at risk of
-death from a variety of causes, including murders, infectious infections, and other health difficulties.(Emily Widra, (2017))
-
-Understanding the interconnectedness of these mortality variables is critical for designing holistic measures to improve
-prisoner safety and well-being((Emily Widra, (2017)). It demands a comprehensive strategy that tackles both immediate concerns, such as
-suicide prevention, as well as the larger structural factors that contribute to higher prison death rates.
-
-Gender Disparities in Prison Health: Unveiling Unique Challenges
-
-In the third portion of our investigation, we examine the subtle disparities in the health experiences of male and female
-convicts. While many health issues are common, gender-specific characteristics play an important role in defining the health
-outcomes of jailed people. Women inmates, for example, frequently confront particular reproductive health challenges,
-as well as an increased risk of sexual assault while in correctional facilities.
-
-Understanding and treating gender differences in prison health is critical for establishing focused therapies that
-meet the unique requirements of both male and female inmates. Recognizing the unique issues experienced by each gender
-group is an important step toward establishing a more equal and compassionate approach to healthcare in the prison system.
-
-Conclusion: A Call for Holistic Reforms
-
-To summarize, the health and safety of inmates today require immediate attention
-and thorough improvements. The complex network of physical health issues, mental health crises,
-mortality risks, and gender imbalances inside the jail system needs a comprehensive strategy.
-Addressing these concerns necessitates not just urgent actions, such as increased healthcare access and mental health
-assistance, but also a larger commitment to rethinking the purpose and consequences of incarceration on individual
-well-being(Ghazanfari, H., & Farokhzadian, J. (2022)).This article intends to add to the continuing discussion about prison health by arguing for policies and
-practices that prioritize the dignity, safety, and general well-being of people housed in correctional institutions.
 

### 2024-02-16 paper/paper.qmd
included the introduction with updated references

@@
 # Introduction
+
+The rising death rates among prisoners in the challenging environment of American jails show discrepancies, difficulties,
+and institutional failings. Issues including institutional prejudice and poor healthcare have been brought to light by
+a recent news report. For instance, the article talks about a study which discovered that although 0.85% of Americans
+with asthma were incarcerated, just 0.15% of asthma patients received treatment, an inequality that was more than five
+times higher (Murez_2023). Our goals are to identify the causes of the growing death rate, recognize the influence of
+demographics and highlight the significant negative effects of prison circumstances on the mental health of prisoners.
+
+According to the US Justice Department's Office of the Inspector General (OIG), 187 individuals died by suicide inside
+Federal Bureau of Prisons(BoP) institutions between 2014 and 2021. The staff providing psychiatric services in the prisons
+stated that these deaths could have been avoided if the facilities followed protocol and provided inmates with the necessary
+resources and treatment. In addition, the study listed 56 "accidental" fatalities and 89 homicides that occurred during that
+time period. It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths.(Levin 2024)
+News reports highlight the critical need for a full understanding of prisoners' problems, going beyond simple statistics.
+This has sparked calls for reform and raised concerns about the efficacy of healthcare in prisons.
+
+Homicides accounted for the second-highest number of fatalities recorded in the survey, with one prisoner, Bulger, being
+beaten to death by other inmates in 2018 (Associated Press_2024). Editorials raise ethical concerns regarding the prison
+system by arguing that inequality in healthcare and the high rate of avoidable diseases among prisoners are reflections
+of larger social issues. Investigative articles highlight patterns of prejudice and neglect while exposing the realities
+of minorities within jails.
+
+As we dig deeper, our focus goes beyond statistics, aiming to amplify the voices and experiences of those directly
+affected. This exploration goes beyond the numbers, delving into the personal dimensions of the prison experience and
+laying the groundwork for a comprehensive analysis that seeks to drive change in a system facing significant challenges.
 

### 2024-02-16 paper/paper.qmd
References updated

@@
 format: pdf
-toc: true
 number-sections: true
@@
 ---
+\newpage
 
+\tableofcontents
+
+\newpage
 ```{r}
@@
 with asthma were incarcerated, just 0.15% of asthma patients received treatment, an inequality that was more than five
-times higher (Murez_2023). Our goals are to identify the causes of the growing death rate, recognize the influence of
+times higher [@Murez_2023]. Our goals are to identify the causes of the growing death rate, recognize the influence of
 demographics and highlight the significant negative effects of prison circumstances on the mental health of prisoners.
@@
 resources and treatment. In addition, the study listed 56 "accidental" fatalities and 89 homicides that occurred during that
-time period. It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths.(Levin 2024)
+time period. It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths [@Levin_2024].
 News reports highlight the critical need for a full understanding of prisoners' problems, going beyond simple statistics.
@@
 Homicides accounted for the second-highest number of fatalities recorded in the survey, with one prisoner, Bulger, being
-beaten to death by other inmates in 2018 (Associated Press_2024). Editorials raise ethical concerns regarding the prison
+beaten to death by other inmates in 2018 [@Associated_Press_2024]. Editorials raise ethical concerns regarding the prison
 system by arguing that inequality in healthcare and the high rate of avoidable diseases among prisoners are reflections

### 2024-02-16 paper/paper.qmd

@@
 
-The rising death rates among prisoners in the challenging environment of American jails show discrepancies, difficulties,
-and institutional failings. Issues including institutional prejudice and poor healthcare have been brought to light by
-a recent news report. For instance, the article talks about a study which discovered that although 0.85% of Americans
-with asthma were incarcerated, just 0.15% of asthma patients received treatment, an inequality that was more than five
-times higher [@Murez_2023]. Our goals are to identify the causes of the growing death rate, recognize the influence of
-demographics and highlight the significant negative effects of prison circumstances on the mental health of prisoners.
+The rising death rates among prisoners in the challenging environment of American jails show discrepancies, difficulties, and institutional failings.
+Issues including institutional prejudice and poor healthcare have been brought to light by a recent news report.
+For instance, the article talks about a study which discovered that although 0.85% of Americans with asthma were incarcerated, just 0.15% of asthma patients received treatment, an inequality that was more than five times higher [@Murez_2023].
+Our goals are to identify the causes of the growing death rate, recognize the influence of demographics and highlight the significant negative effects of prison circumstances on the mental health of prisoners.
 
-According to the US Justice Department's Office of the Inspector General (OIG), 187 individuals died by suicide inside
-Federal Bureau of Prisons(BoP) institutions between 2014 and 2021. The staff providing psychiatric services in the prisons
-stated that these deaths could have been avoided if the facilities followed protocol and provided inmates with the necessary
-resources and treatment. In addition, the study listed 56 "accidental" fatalities and 89 homicides that occurred during that
-time period. It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths [@Levin_2024].
+According to the US Justice Department's Office of the Inspector General (OIG), 187 individuals died by suicide inside Federal Bureau of Prisons(BoP) institutions between 2014 and 2021.
+The staff providing psychiatric services in the prisons stated that these deaths could have been avoided if the facilities followed protocol and provided inmates with the necessary resources and treatment.
+In addition, the study listed 56 "accidental" fatalities and 89 homicides that occurred during that time period.
+It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths [@Levin_2024].
+
 News reports highlight the critical need for a full understanding of prisoners' problems, going beyond simple statistics.
@@
 
-Homicides accounted for the second-highest number of fatalities recorded in the survey, with one prisoner, Bulger, being
-beaten to death by other inmates in 2018 [@Associated_Press_2024]. Editorials raise ethical concerns regarding the prison
-system by arguing that inequality in healthcare and the high rate of avoidable diseases among prisoners are reflections
-of larger social issues. Investigative articles highlight patterns of prejudice and neglect while exposing the realities
-of minorities within jails.
+Homicides accounted for the second-highest number of fatalities recorded in the survey, with one prisoner, Bulger, being beaten to death by other inmates in 2018 [@Associated_Press_2024].
+Editorials raise ethical concerns regarding the prison system by arguing that inequality in healthcare and the high rate of avoidable diseases among prisoners are reflections of larger social issues.
+Investigative articles highlight patterns of prejudice and neglect while exposing the realities of minorities within jails.
 
-As we dig deeper, our focus goes beyond statistics, aiming to amplify the voices and experiences of those directly
-affected. This exploration goes beyond the numbers, delving into the personal dimensions of the prison experience and
-laying the groundwork for a comprehensive analysis that seeks to drive change in a system facing significant challenges.
+As we dig deeper, our focus goes beyond statistics, aiming to uncover trends of causes of prisoner death rates in the United States and examine these death rates both from a comprehensive standpoint and delve into demographic breakdowns, specifically considering gender and race.
+We also draw in Covid-19 data about new cases to support our analysis of wellbeing in prison.
+
+The paper is further organised into three sections: Data, Results, and Discussion.
+In the Data section, I discuss how the datasets to be used for the analysis were obtained, pre-processed and cleaned.
+I will also explain the variables of interest in the datasets for my analysis.
+The Results section will then highlight and discuss the trends and associations found during the analysis.
+Lastly, the Discussion section will talk about some interesting trends found in Results in depth, link it to the real world and also highlight the weaknesses of our analysis.
+
+\newpage
 
@@
 
-- x1 (The race of the prisoner)
+- x1 (the race of the prisoner)
 - Cause (the cause of death of the prisoner)
@@
 
-- x1 (The gender of the prisoner)
+- x1 (the gender of the prisoner)
 - Cause (the cause of death of the prisoner)
@@
 
-- Date (The date of the new case)
-- New_cases_per_100k (The number of new cases recorded on the specific date per 100k)
-- Group (The group of people the case belonged to)
+- Date (the date of the new case)
+- New_cases_per_100k (the number of new cases recorded on the specific date per 100k)
+- Group (the group of people the case belonged to)
 
@@
 The other causes have been almost the same with a few spikes over the years.
+
+\newpage
 
@@
 
+\newpage
+
 ```{r}
@@
 
-## Weaknesses and next steps
+## Weaknesses
+While our analysis of prisoner mortality provides valuable insights, it is important to acknowledge certain limitations.
+The binary representation of gender in the original paper oversimplifies a complex spectrum, neglecting nuances beyond the male-female binary.
+
+Additionally, the absence of data beyond 2016 restricts our ability to capture recent developments, and the lack of detailed information about other races limits a comprehensive understanding of racial disparities.
+
+Also, the datasets’ focus on co-ed adult prisons overlooks the diverse landscape of correctional facilities, including juvenile and gender-specific institutions, which could yield distinct insights.
+
+Furthermore, the dataset gives collective information about the prisons in the US.
+The inclusion of regional split information could help us uncover regional variations in behavior and pre-existing medical conditions to name a few factors.
+
+Lastly, exploring the intersectionality of race and gender would provide a more nuanced understanding of the complex factors influencing prisoner mortality.
+
+## Next Steps
+To take our study one step beyond, we could look for datasets that would fulfill the weaknesses of our datasets and combine that with our existing data.
+This could strengthen the general investigation that we are trying to make.
+We could also explore the data of other countries with either similar geography or demographic to compare how each of the prison systems perform under the general well-being category.
 

### 2024-02-16 paper/paper.qmd
discussion added

@@
 # Discussion
+In this paper, the issue of mortality rates among U.S. inmates is analyzed and conducted.
+There are multiple datasets that have been utilized to investigate demographic differences, causes of death,
+and the potential impact of prison conditions. We also looked at the impact of the COVID-19 pandemic.
+The study digs deeper into a complex analysis of mortality trends, drawing from datasets that include information
+on gender, race, and specific causes of death. We have explored three different categories of the issue with the
+wellbeing of prisoners.
+
+The first is the different causes of death in the incarcerated in the categories of gender(Male/Female).
+We found Cancer to cause most of the deaths in Male and Female prisoners. Heart disease was almost 0.25 times more
+prevalent in Male prisoners than in Female prisoners. A significant difference is found in the rates of suicide between
+the two genders. Female prisoners are more likely to commit suicide. This indicates a huge concern in the community of
+prisoners. It shadows upon the mental health resources that are not provided to those in need. Men are more prone to
+having disputes in the cell and hence more likely to see deaths due to homicides[@Levin_2024].
+Our dataset from 10 years ago still supports this argument that is being discussed today.
+
+Secondly, we analyzed a data set that shows us the different causes of death in prisoners
+categorized by race. We found that the prisoners of the black community were a lot more affected by aids than any other
+race. This also directly coincides with the fact today that incarcerated individuals with HIV represented about 2.2%
+of the U.S. total, but got only 0.73% of HIV antivirals, a threefold difference, researchers estimated [@Murez_2023].
+Expanding on the suicide rates in prisons from this dataset we also noticed that the prisoners from other races have
+lost their lives to suicide more than black, hispanic and white prisoners. We have seen this ongoing trend in
+today's society as well where with depression, estimated prevalence was 15% in inmates compared to 7.64% in others.
+Severe mental illness was 13.12% compared to 4.89% in others [@Murez_2023]. We also see from this dataset and our
+analysis that hispanic prisoners have way more liver disease related deaths than any other race. The lack of health care in prisons is affecting the people inside and out at an alarming rate.
+
+Our third data set shadows upon the negative impacts that COVID-19 had upon the prisoners in America. Due to the
+conditions of the prisons and the lack of care for the inmates, from our dataset we analyzed that the number of cases
+in prisoners were significantly higher than the number of cases outside prison. Despite considering the lack of
+healthcare and sanitary methods that should have been implemented in these isolated environments, these numbers were
+shocking. During the most difficult time of Covid-19 where the highest number of people were affected, if 70 people
+were affected outside, almost 100 people were affected inside these prisons. In a statistical study done in 2022,
+it was concluded that from the end of February 2020 to the end of February 2021, a total of 196 correctional staff
+in state and federal prisons died as a result of COVID-19[@NadelE.AnnCarson_2022].
 

### 2024-02-16 paper/paper.qmd
discussion finished

@@
 
+In conclusion, the findings illuminate an alarming increase in prisoner mortality rates, particularly attributed to cancer and heart
+diseases. The persistent rise raises concerns about the overall well-being of incarcerated individuals. The vulnerability
+ of inmates to COVID-19, as evidenced by the significantly higher infection rates among prisoners, highlights the
+ urgent need for enhanced preventive measures and healthcare within correctional facilities. The study
+ uncovers pronounced gender and racial differences in prison mortality. Males exhibit higher mortality
+ rates across various causes, emphasizing the need for gender-specific healthcare strategies. Racial differences
+ reveal distinct health challenges faced by different groups, necessitating tailored interventions.  Exploring
+ the influence of specific prison conditions, such as access to medical care and systemic biases, can provide deeper
+ insights. Comparative studies with international datasets may offer a broader perspective on correctional healthcare.
+ In all, our paper aims to strengthen the argument that the lack of attention provided to people in correctional
+ facilities trying to improve their lives has reached an abysmal state. Instead of paying attention to these concerns
+ in society, prisoners are often looked down upon and neglected. Our paper firmly portrays the basic fact that as humans,
+even prisoners have rights to access basic healthcare and clean living conditions that will help them in order to grow
+into better individuals in the right environment.
+
 ## Weaknesses

### 2024-02-16 paper/paper.qmd
Final changes

@@
   - Aamishi Avarsekar
-thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
+thanks: "Code and data are available at: [https://github.com/kaavyakalani26/prison_health.git](https://github.com/kaavyakalani26/prison_health.git) and replication aspects are available at: [https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index](https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index)"
 date: today
@@
 
+\newpage
+
 # Discussion
+
+## Learnings and Inferences
+
 In this paper, the issue of mortality rates among U.S. inmates is analyzed and conducted.
